Automated Verification for Real-Time Systems using Implicit Clocks and an Extended Antimirov Algorithm

### 2022-04-08 paper/tv.tex
m

\begin{definition}[System configuration]
A system configuration is a pair $c = (V , e)$ where $V$ is a variable valuation function and $e$ is an expression.
\end{definition}

@@

{{\small
\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   Wait                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  t {\leq} d 
}{(V, \textbf{delay}[d]) \xrightarrow[]{\text{t}} (V, \textbf{delay} [d \text{-}t])}\ [de_1]} 
\qquad \qquad 
\code{\frac{
}{(V, \textbf{delay} [0]) \xrightarrow[]{\tau} (V, ())}\ [de_2]} 
\end{flalign*}}}

The above captures behaviors of process \code{\textbf{delay}[d]}. 
Rule \code{de_1} states that the process may idle for 
any amount of time as long as it is less than or equal to \codem{d} time units; 
Rule \code{de_2} states that the process terminates immediately after 
\codem{d} becomes \codem{0}.
{{\small\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   timeout                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e_1) \xrightarrow[]{\anyevent{A}} (V^\prime,e_1^\prime)
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\anyevent{A}} (V^\prime,e^\prime)}\ [to_1]} 
\quad  
\code{\frac{
  (V,e_1) \xrightarrow[]{\tau} (V^\prime,e_1^\prime)  
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (V^\prime, e_1^\prime\ \textbf{timeout}[d]\ e_2 )}\ [to_2]} 
\\
\code{\frac{
  (V,e_1) \xrightarrow[]{\text{t}} (V,e_1^\prime)  \quad (t {\leq} d) 
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (V,e_1^\prime\ \textbf{timeout} \  [d\text{-}t] \ e_2)}\ [to_3]} 
\quad  
\code{\frac{
}{(V, e_1\ \textbf{timeout}   [0] \ e_2) \xrightarrow[]{\tau} (V, e_2)}\ [to_4]} 
\end{flalign*}}}
If an observable event \anyevent{A} can be engaged by \code{e_1}, 
then \code{e_1\ \textbf{timeout}[d]\ e_2} becomes \code{e_1^\prime} (rule \code{[to_1]}). 
An invisible transition does 
not solve the choice (rule \code{[to_2]}). 
If \code{e_1} may idle for less than or equal 
to \codem{d} time units, so is the composition (rule \code{[to_3]}). 
When \codem{d} becomes \codem{0}, 
\code{e_2} takes over control by a silent transition (rule \code{[to_4]}).
{{\small\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   deadline                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e) \xrightarrow[]{\anyevent{A}} (V^\prime,e^\prime)
}{(V, \textbf{deadline} [d] \ e ) \xrightarrow[]{\anyevent{A}} (V^\prime, \textbf{deadline}  [d] \ e^\prime)} [dl_1]} 
\quad 
\code{\frac{
  (V,e) \xrightarrow[]{\text{t}} (V,e^\prime)  \quad (t {\leq} d)
}{(V, \textbf{deadline}  [d]\ e ) \xrightarrow[]{\text{t}} (V, \textbf{deadline}  [d \text{-}t] \ e^\prime )} [dl_2]} 
\end{flalign*}}}

@@

{{\small\begin{flalign*}
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  %%   deadline                 %%
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \code{\frac{
    V \not\models  b 
  }{(V, [b] \ e ) \xrightarrow[]{t} (V, [b] \ e)}\ [gu_1]} 
  \ \ \qquad
  \code{\frac{
    V \models  b 
  }{(V, [b] \ e ) \xrightarrow[]{\tau} (V, e)}\ [gu_2]} 
  \ \ \qquad 
  \code{\frac{
  }{(V, \alpha ) \xrightarrow[]{\tau} (\alpha(V), ())}\ [assign]} 
\end{flalign*}}}

{{\small\begin{flalign*}
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  %%   deadline                 %%
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \code{\frac{
    (V,e_1) \xrightarrow[]{\anyevent{A}} (V^\prime,e_1^\prime) 
  }{(V, e_1 || e_2 ) \xrightarrow[]{\anyevent{A}} (V^\prime, e_1^\prime || e_2)}\ [pa_1]} 
  \ \   
  \code{\frac{
    (V,e_2) \xrightarrow[]{\anyevent{A}} (V^\prime,e_2^\prime) 
  }{(V, e_1 || e_2 ) \xrightarrow[]{\anyevent{A}} (V^\prime, e_1 || e_2^\prime)}\ [pa_2]} 
\ \ 
\code{\frac{
  (V,e_1) \xrightarrow[]{\anyevent{A}/\tau} (V^\prime,e_1^\prime) 
}{(V, e_1; e_2 ) \xrightarrow[]{\anyevent{A}/\tau} (V^\prime, e_1^\prime ; e_2)}\ [se]} 
\end{flalign*}}}

@@

Now, the inclusions are disjunction-free formulas. 
Next we provide the definitions and the key implementations\footnote{As 
the implementations according to basic regular expressions can be found 
in prior work \cite{keil2014symbolic}. 
%Here, we focus on presenting 
%the definitions and how do we deal with dependent values in the effects, 
%as the key novelties of this work. 
%More detail implementations can be found in the extended technical report\citeref{Demo}.
} of Nullable, First and Derivative respectively. Intuitively, 
the Nullable function \code{\delta_{{\pi}}(\es)} returns a boolean 
value indicating whether \code{{\pi} \wedge \es} contains the empty 
trace; the First function \code{fst_{{\pi}}( \es)} computes a set of 
 initial \emph{heads}, denoted as \code{h},  of \code{{\pi} \wedge \es}; and the Derivative 
function \code{D_{h}^{{\pi}}(\es)} computes a next-state 
effects after eliminating the head  \code{h} from the current 
effects \code{{\pi} \wedge \es}.

@@

\begin{definition}[Timed Event Subsumption]\label{Subsumption}
Given two instances \code{I} and \code{J}, we define the subset relation \code{I {\subseteq} J} as: the set of present signals in \code{J} is a subset of the set of present signals in \code{I}, and the set of absent signals in \code{J} is a subset of the set of absent signals in \code{I}\footnote{As in having more constraints refers to a smaller set of satisfying instances.}.
Formally, 
 \begin{gather*} 
 \code{I {\subseteq} J \Leftrightarrow  \{\textbf{S}\ |\ (\textbf{S} {\mapsto} Present) \in J \}  \subseteq  \{\textbf{S}\ |\ (\textbf{S} {\mapsto} Present) \in I \}} \\
  \code{ \quad\  and  \ 
 \{\textbf{S}\ |\ (\textbf{S} \mapsto Absent) \in J \}  \subseteq  \{\textbf{S}\ |\ (\textbf{S} \mapsto Absent) \in I \}
 }
\end{gather*}

\end{definition}

\begin{definition}[Time Instances Subsumption]\label{Subsumption1}
Given two time instances \code{(I, \pi_1, t_1)} and \code{(J, \pi_2, t_2)}, we define the subset relation \code{(I, \pi_1, t_1) {\subseteq} (J, \pi_2, t_2)} as:  \code{I {\subseteq} J} and \code{\pi_1[t_2/t_1]  {\Rightarrow}   \pi_2  }.

\end{definition}

\begin{definition}[Partial Derivative\footnote{Intuitively, the partial derivative refers to the left quotient of a language equation, for example, for REs, \code{x^{{-}1} \llbracket x\cdot y\rrbracket {=} y};  \code{y^{{-}1} \llbracket x\cdot y\rrbracket {=} \bott}; and  \code{y^{{-}1} \llbracket x + y\rrbracket {=} \epsilon}. Here we come up with a new notion of partial derivative for \timedEffects. 
}]\label{Derivative}
The partial derivative \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect)} 
of effects \code{\pi : \effect} 
w.r.t. a event \code{(I, \Pi_1, t^\prime)} computes the  effects for the left quotient \code{(I, \Pi_1, t^\prime)^{\text{-}1}\llbracket \pi : \effect \rrbracket}. \footnote{For understanding of the derivative rule for \code{\effect\mysharp  t} (cf. \tabref{tab:rewriting_tree_send}  step \textcircled{3}), it is essentially splitting \code{\effect} with a head and a tail, bound with fresh terms \code{t_1} and \code{t_2} respectively. Meanwhile it unifies \code{t_1} with \code{t^\prime} and  adds the constraint \code{t_1 {+} t_2 {=} t}.}
%\footnote{As the reader may notice, our First and Derivative function does not handle the timed sequences constructed by \code{||}, as the synchronous parallel is reduced during the normalization phase.}
\begin{gather*}
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\bott) {=}}  
 \code{False : \bott} 
\qquad\ 
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \empt) {=}}   
\code{False : \bott} 
\qquad \ 
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_1 ||\effect_2)} {=} \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_1)} || \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_2)} 
    \\
        \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  {\effect^\star}) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  {\effect}) \cdot ({\effect^\star})}  
        \qquad\qquad
\\
{
 \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(   \effect_1 \seq \effect_2) {=} }
\begin{cases}
      \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(   \effect_1) \cdot  \effect_2 \vee  D^{\pi}_{(I, \Pi_1, t^\prime)}( \effect_2)} \quad &\code{if\ \delta(\effect_1) {=} true}   \\
      \code{D^{\pi}_{(I, \Pi_1, t^\prime)}  ( \effect_1) \cdot  \effect_2} & \code{if\ \delta(\effect_1) {=} false} 
    \end{cases} 
    }
 \\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1 \choice \effect_2) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1) \choice D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_2)} 
%\\
%\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1 \ ||\  \effect_2) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1) \ ||\ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_2)} 
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect \mysharp  t) } {=}
\code{
\pi \wedge(t_1{+} t_2 {=}t) \wedge (t_1{=}t^\prime) : (D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect))\mysharp  t_2} \ \ \ 
\code{(fresh \ t_1\ t_2 )}
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\textbf{S}?)} {=} \code{\pi: \empt\ (}\code{ I {\subseteq} \{\textbf{S}\})}
\qquad
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\textbf{S}?)} {=} \code{\pi: \textbf{S}?\ (}\code{ I {\not\subseteq} \{\textbf{S}\})}
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(J)} {=} \code{\pi: \empt\ (}\code{ I {\subseteq} J)}
\qquad\ 
    \code{D^{\pi}_{(I, \Pi_1, t^\prime)}( J)} {=} \code{False : \bott \ (}\code{I {\not\subseteq} J)} 
\end{gather*}
\end{definition}

>>>>

\begin{definition}[System configuration]
A system configuration is a pair $c = (\env , e)$ where $\env$ is a variable valuation function and $e$ is an expression.
\end{definition}

@@

{{\small
\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   Wait                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  t {\leq} d 
}{(\env, \textbf{delay}[d]) \xrightarrow[]{\text{t}} (\env, \textbf{delay} [d \text{-}t])}\ [de_1]} 
\qquad \qquad 
\code{\frac{
}{(\env, \textbf{delay} [0]) \xrightarrow[]{\tau} (\env, ())}\ [de_2]} 
\end{flalign*}}}

The above captures behaviors of process \code{\textbf{delay}[d]}. 
Rule \code{de_1} states that the process may idle for 
any amount of time as long as it is less than or equal to \codem{d} time units; 
Rule \code{de_2} states that the process terminates immediately after 
\codem{d} becomes \codem{0}.
{{\small\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   timeout                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (\env,e_1) \xrightarrow[]{\anyevent{A}} (\env^\prime,e_1^\prime)
}{(\env, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\anyevent{A}} (\env^\prime,e^\prime)}\ [to_1]} 
\quad  
\code{\frac{
  (\env,e_1) \xrightarrow[]{\tau} (\env^\prime,e_1^\prime)  
}{(\env, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (\env^\prime, e_1^\prime\ \textbf{timeout}[d]\ e_2 )}\ [to_2]} 
\\
\code{\frac{
  (\env,e_1) \xrightarrow[]{\text{t}} (\env,e_1^\prime)  \quad (t {\leq} d) 
}{(\env, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (\env,e_1^\prime\ \textbf{timeout} \  [d\text{-}t] \ e_2)}\ [to_3]} 
\quad  
\code{\frac{
}{(\env, e_1\ \textbf{timeout}   [0] \ e_2) \xrightarrow[]{\tau} (\env, e_2)}\ [to_4]} 
\end{flalign*}}}
If an observable event \anyevent{A} can be engaged by \code{e_1}, 
then \code{e_1\ \textbf{timeout}[d]\ e_2} becomes \code{e_1^\prime} (rule \code{[to_1]}). 
An invisible transition does 
not solve the choice (rule \code{[to_2]}). 
If \code{e_1} may idle for less than or equal 
to \codem{d} time units, so is the composition (rule \code{[to_3]}). 
When \codem{d} becomes \codem{0}, 
\code{e_2} takes over control by a silent transition (rule \code{[to_4]}).
{{\small\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   deadline                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (\env,e) \xrightarrow[]{\anyevent{A}} (\env^\prime,e^\prime)
}{(\env, \textbf{deadline} [d] \ e ) \xrightarrow[]{\anyevent{A}} (\env^\prime, \textbf{deadline}  [d] \ e^\prime)} [dl_1]} 
\quad 
\code{\frac{
  (\env,e) \xrightarrow[]{\text{t}} (\env,e^\prime)  \quad (t {\leq} d)
}{(\env, \textbf{deadline}  [d]\ e ) \xrightarrow[]{\text{t}} (\env, \textbf{deadline}  [d \text{-}t] \ e^\prime )} [dl_2]} 
\end{flalign*}}}

@@

{{\small\begin{flalign*}
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  %%   deadline                 %%
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \code{\frac{
    \env \not\models  b 
  }{(\env, [b] \ e ) \xrightarrow[]{t} (\env, [b] \ e)}\ [gu_1]} 
  \ \ \qquad
  \code{\frac{
    \env \models  b 
  }{(\env, [b] \ e ) \xrightarrow[]{\tau} (\env, e)}\ [gu_2]} 
  \ \ \qquad 
  \code{\frac{
  }{(\env, \alpha ) \xrightarrow[]{\tau} (\alpha(\env), ())}\ [assign]} 
\end{flalign*}}}

{{\small\begin{flalign*}
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  %%   deadline                 %%
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \code{\frac{
    (\env,e_1) \xrightarrow[]{\anyevent{A}} (\env^\prime,e_1^\prime) 
  }{(\env, e_1 || e_2 ) \xrightarrow[]{\anyevent{A}} (\env^\prime, e_1^\prime || e_2)}\ [pa_1]} 
  \ \   
  \code{\frac{
    (\env,e_2) \xrightarrow[]{\anyevent{A}} (\env^\prime,e_2^\prime) 
  }{(\env, e_1 || e_2 ) \xrightarrow[]{\anyevent{A}} (\env^\prime, e_1 || e_2^\prime)}\ [pa_2]} 
\ \ 
\code{\frac{
  (\env,e_1) \xrightarrow[]{\anyevent{A}/\tau} (\env^\prime,e_1^\prime) 
}{(\env, e_1; e_2 ) \xrightarrow[]{\anyevent{A}/\tau} (\env^\prime, e_1^\prime ; e_2)}\ [se]} 
\end{flalign*}}}

@@

Now, the inclusions are disjunction-free formulas. 
Next we provide the definitions and the key implementations\footnote{As 
the implementations according to basic regular expressions can be found 
in prior work \cite{keil2014symbolic}. 
%Here, we focus on presenting 
%the definitions and how do we deal with dependent values in the effects, 
%as the key novelties of this work. 
%More detail implementations can be found in the extended technical report\citeref{Demo}.
} of Nullable, First and Derivative respectively. Intuitively, 
the Nullable function \code{\delta_{{\pi}}(\es)} returns a boolean 
value indicating whether \code{{\pi} \wedge \es} contains the empty 
trace; the First function \code{fst_{{\pi}}( \es)} computes a set of 
 initial \emph{heads}, denoted as \code{h},  of \code{{\pi} \wedge \es}; and the Derivative 
function \code{D^{\env}_{h}(\pi \wedge \es)} computes a next-state 
effects after eliminating the head  \code{h} from the current 
effects \code{{\pi} \wedge \es}, w.r.t \code{\env}.

@@

\begin{comment}
  \begin{definition}[Heads Subsumption]\label{Subsumption1}
Given two time instances \code{(I, \pi_1, t_1)} and \code{(J, \pi_2, t_2)}, we define the subset relation \code{(I, \pi_1, t_1) {\subseteq} (J, \pi_2, t_2)} as:  \code{I {\subseteq} J} and \code{\pi_1[t_2/t_1]  {\Rightarrow}   \pi_2  }.

\end{definition}
\end{comment}

\begin{definition}[Partial Derivative\footnote{Intuitively, the partial derivative refers to the left quotient of a language equation, for example, for REs, \code{x^{{-}1} \llbracket x\cdot y\rrbracket {=} y};  \code{y^{{-}1} \llbracket x\cdot y\rrbracket {=} \bott}; and  \code{y^{{-}1} \llbracket x + y\rrbracket {=} \epsilon}. Here we come up with a new notion of partial derivative for \timedEffects. 
}]\label{Derivative}
Given any \code{\effect {=} \pi \wedge \es}, and a variable valuation function \code{\env}, 
the partial derivative \code{D^{\env}_{h}(\pi \wedge \es)} 
 computes the  effects for the 
left quotient \code{(h)^{\text{-}1}\llbracket \pi \wedge \es 
\rrbracket_{\env}}. 
%\footnote{As the reader may notice, our First and Derivative function does not handle the timed sequences constructed by \code{||}, as the synchronous parallel is reduced during the normalization phase.}
\begin{gather*}
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\bott) {=}}  
 \code{False : \bott} 
\qquad\ 
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \empt) {=}}   
\code{False : \bott} 
\qquad \ 
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_1 ||\effect_2)} {=} \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_1)} || \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect_2)} 
    \\
        \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  {\effect^\star}) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  {\effect}) \cdot ({\effect^\star})}  
        \qquad\qquad
\\
{
 \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(   \effect_1 \seq \effect_2) {=} }
\begin{cases}
      \code{D^{\pi}_{(I, \Pi_1, t^\prime)}(   \effect_1) \cdot  \effect_2 \vee  D^{\pi}_{(I, \Pi_1, t^\prime)}( \effect_2)} \quad &\code{if\ \delta(\effect_1) {=} true}   \\
      \code{D^{\pi}_{(I, \Pi_1, t^\prime)}  ( \effect_1) \cdot  \effect_2} & \code{if\ \delta(\effect_1) {=} false} 
    \end{cases} 
    }
 \\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1 \choice \effect_2) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1) \choice D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_2)} 
%\\
%\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1 \ ||\  \effect_2) } {=} \code{ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_1) \ ||\ D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect_2)} 
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(  \effect \mysharp  t) } {=}
\code{
\pi \wedge(t_1{+} t_2 {=}t) \wedge (t_1{=}t^\prime) : (D^{\pi}_{(I, \Pi_1, t^\prime)}(\effect))\mysharp  t_2} \ \ \ 
\code{(fresh \ t_1\ t_2 )}
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\textbf{S}?)} {=} \code{\pi: \empt\ (}\code{ I {\subseteq} \{\textbf{S}\})}
\qquad
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(\textbf{S}?)} {=} \code{\pi: \textbf{S}?\ (}\code{ I {\not\subseteq} \{\textbf{S}\})}
\\
\code{D^{\pi}_{(I, \Pi_1, t^\prime)}(J)} {=} \code{\pi: \empt\ (}\code{ I {\subseteq} J)}
\qquad\ 
    \code{D^{\pi}_{(I, \Pi_1, t^\prime)}( J)} {=} \code{False : \bott \ (}\code{I {\not\subseteq} J)} 
\end{gather*}
\end{definition}
